A Computational Account of Default-Mode Function\\ by Control Theory and Reinforcement Learning

### 2016-10-07 paper/paper.tex
RL-semantics part 1

The vmPFC subserves less ambiguous subjective-value-related evaluative processes
reward-informed and risk estimation of self-relevant environmental stimuli.
This DMN node is more closely associated with
orchestrating adapted behavior by bottom-up-driven
processing of “what matters now”,
probably drawing on model-based value representations
\cite{doherty2015structure}.
Quantative lesion findings across 344 human individuals confirmed
a gross impairment in value-based decision making
(Glaescher et al., 2016).
The vmPFC is preferentially connected with limbic and reward-related areas.
The vmPFC has been observed to have monosynaptical connections
with the nucleus accumbens
in axonal tracing studies in monkeys (Haber et al., 1995; Ferry et al., 2000).
The NAc is thought to be linked to reward mechanisms that not
only modulate motivated behavior towards basic survival needs but also
subserve reinforcement learning in humans more broadly
(O'Doherty 2015).
This is consistent with diffusion MRI (dMRI) tractography in humans and monkeys
(Croxson et al., 2005) that
quantified this brain regions to
be substantially more likely connected to the vmPFC than dmPFC in both species.
Two functional connectivity modalities in humans strongly connected
the vmPFC with the nucleus accumbens, hippocampus (HC),
and posteriomedial cortex.
%
The vmPFC is often proposed to be involved in (external) emotional
reactions and own (visceral) arousal.
Real or imaginged bodily states could be mapped in the vmPFC
as a bioregulatory disposition governing cognition
and decision making (Damasio, 1996; Nauta, 1971)
In neuroeconomic studies of human decision making,
the vmPFC consistently reflects an individual’s subjective
valuation (Behrens et al., 2008; Boorman et al., 2009; Knutson et al., 2005; Plassmann et al., 2007)
This may be why performance within and across participants
was related to state encoding in OFC.
Such a “cognitive map” of action space was argued to encode
the current task state even when states are unobservable from sensory input,
which was shown to be critical for behavior.
The GMV of the vmPFC and nucleus accumbens
correlated with indices of social reward attitudes and
value-guided behavior (Lebreton et al., 2009).
Additionally,
independent whole-brain analyses from structural
neuroimaging studies related the gray-matter volume GMV of the vmPFC,
more consistently than any other
brain region, to indices of
social competence and social network complexity,
among the most complicated decision that humans and monkeys take
(Lebreton et al., 2009; Powell et al., 2010; Lewis et al., 2011; Sallet et al., 2011).
%

@@

By construction, in general, maximizing the expectation in  \eqref{eq:qval} is intractable due to
the enormous number of possible trajectories the agent might need to
consider.
(If the Bellman-based computational ideas have a corresponding
neurobiological implementation, this provides a framework of how
optimal behavior could be computable in the human brain.)
One popular simplification that removes this issue is to only consider the sub-family of
\textit{deterministic policies}, which map each state to the best single best action which
should be taken at thin state, and therefore define functions from states to actions.
\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The celebrated \textit{Bellman's equation} [Sutton and Barto 1998] then takes the simple form
\begin{equation}
  Q^\pi(s, a) = \mathbb E_{s' \sim \rho(.|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q^\pi(s', a')]
  = \mathbb E_{s' \sim \rho(.|s,a)} [r(s,a) + \gamma Q^\pi(s', \pi(s'))]
  \label{eq:bellman}
\end{equation}
which provides a recursive decomposition of optimal behavior.
A complicated dynamic programming optimization can be
broken into simpler subproblems at different time points.
Using the Bellman equation, each state can be associated with a certain value,
to guide action (i.e., improve the policy) towards a better state.
It is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

@@

\section{Relation to other cognitive models (Danilo)}
\paragraph{Predictive Coding and Bayesian Brain Hypothesis}
The predictive coding and Bayesian brain framework 
is a frequently evoked idea related to the default mode function
\cite{bar2007}.
According to this view,
cortical responses emerge from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
This dynamic intercourse enables inference on the world by reconciling
gaps between fresh sensory input and stored prior information. 
Feed-forward sensory experience is constantly calibrated by
top-down modulation at various hierarchical processing levels.
back projections probably outnumber by far the existing input projection in the brain
\cite{salin1995corticocortical}.
It can explains the constructive, generative nature of sensory perception
\cite{friston2010free} and
why motor action is intimately related to sensory expectations
\cite{wolpert1995internal}.
At each stage of neural processing,
an internally generated prediction of the sensory input is
directly compared against the actual input from the environment.
A prediction error at one of the processing levels,
i.e. the difference between sensory input
and internally predicted sensation,
incurs plasticity changes of neuronal back-projections (i.e., model parameters)
for gradually improved future prediction of the environment.
Contextual integration is maintained by top-down modulation of sensorimotor
processing by context-specific a-priori information.
The generative model of how data arise in the
world hence be incorporated into
the current neuronal wiring.
Indeed,
This process permits updates of the internal model of the environment
to best fit the incoming sensory examples.

@@

\paragraph{Semantic Hypothesis}

The proposal promoted here is that the brain is continually engaged in generating predictions, and that these predictions rely on associative activation. 
what people do when their mental capacity is not completely consumed by a specific task is to generate associations
Traditionally this process has been considered as recognition, classification, or even a type of memory retrie- val, but in the present context I treat this process as analogy- making instead. 
in analogy, the emphasis is on ‘what is it like?
Analogy- based mappings of properties manifest themselves in processes ranging from perception and memory [17] to stereotypic judgments and prejudice
While our existing memories are used to derive analo- gies and activate predictions, they are constantly being updated. The analogical process, in addition to affording the interpretation of our environment, subsequently aug- ments previous representations in a way that fosters increasingly flexible future analogies.
(Bar)

phantastic organ (from Greek phantastikos, the ability to create mental images)
Friston2014

Navigation processes require associative processing to link land- marks with spatial locations, episodic memories require associa- tions for binding the constituents of an episode, emotional processing and the concomitant heightened arousal they entail are tightly associated with affective memories, and our percep- tion of ourselves and others require associations formed with experience to help interpret the present and anticipate the future.
In some respects, predictions may be considered a higher level of thought-element than associations; they also provide the basis to the many processes detailed above as occupying the unconstrained mind (e.g., planning, navigating, anticipating the thoughts and actions of others, stereotyping, imagining our- selves in the future, affective forecasting, etc.). The brain seems to be able to extract regularities and familiar patterns from our environment, and associate them with corresponding knowl- edge in memory to afford the generation of predictions. This idea of predictions as the basis of many, if not most, cognitive processes in the human brain will have to be developed sepa- rately because the observations we described here cannot speak to it directly
Bar2007

as showing that animals could learn a “cognitive map of the environment” in the absence of rewards or penalties, and that they could use the map later when they were motivated to reach a goal (Tolman, 1948).

this perspective can be traced back to Hermann von Helmholtz and the notion of unconscious inference: that is, a pre-rational mechanism by which visual impressions are formed (eg, the seemingly automatic but erroneous belief that the sun rises and sets in the sky, as opposed to the truth that the Earth rotates around it).

mem- ory-based predictions, and these predictions are generated continually either based on gist information gleaned from the senses or driven by thought. The emphasis in this.
The proposed account integrates three primary components. The first is associations, which are formed by a lifetime of extracting repeating patterns and statisti- cal regularities from our environment, and storing them in memory. The second is the concept of analogies, whereby
Corresponding author: Bar, M. (bar@nmr.mgh.harvard.edu). Available online 4 June 2007.
we seek correspondence between a novel input and existing representations in memory (e.g. ‘what does this look like?’). Finally, these analogies activate associated representa- tions that translate into predictions 
(Bar)

A second important functional account of the DMN is the 'semantic hypothesis'. It holds that retrieval of building blocks of world knowledge enables language and other high-level faculties. The DMN central involvement in semantic processing results in its implication in many a-priori unrelated aspects of human behavior, such as reasoning, planning, problem solving and decision making (Binder, Desai, Graves, \& Conant, 2009; Schilbach et al., 2008; Suddendorf \& Corballis, 2007). This is also illustrated by the substantial relationship between semantic knowledge and words (Binder et al., 2009). The engagement of DMN in language processing matches the observation that the large majority of the DMN regions such as TPJ receive no direct input from primary sensory areas acting more as a “hermeneutic hub” (Binder et al., 2009; Seghier, 2013). The DMN activation rather enables the organization and processing of internally generated information in absence of stimulus presentation (Binder et al., 1999). The proposed role of the DMN in semantic processing is further highlighted by the convergence of these regions in a reciprocal relationship as an interface between the allocentric (bird’s eye view) and egocentric (first person view) perspective; in this way self-reflection, social knowledge, and autobiographical memories are integrated. Moreover, the semantic account exquisitely explains the engagement of DMN in both recalling past and envisioning future (Demblon, Bahri, \& D'Argembeau, 2016; Hassabis, Kumaran, Vann, \& Maguire, 2007; Schacter et al., 2007; Spreng, Mar, \& Kim, 2009), in that the internally stored in language world knowledge may lay the contextual foundations for envisioning future events as well as memory processes (Bzdok et al., 2012). Hence, the reported retrograde amnesia and the simultaneous impairment of imagining novel experience across patients with hippocampus lesions ties in well with this model (Hassabis \& Maguire, 2007). In other words, DMN activation might underlie a time dimension independent cognitive process of constructing contextual scenarios out of detached memories (Buckner \& Carroll, 2007; Gaesser, Spreng, McLelland, Addis, \& Schacter, 2013; Hassabis \& Maguire, 2007). Summing up the 'semantic account', DMN offers the conceptual framework that underlies not only language but also other major high-level functions of human behavior.

\begin{itemize}
  \item The notion of semantic or knowledge association is
  incorporated into the MDP as a function of the state and action
  history.
  \item Both: there is no evidence to indicate that predictions of various levels of complexity, abstraction, timescale and purpose use mech- anisms that are qualitatively different
\end{itemize}

\paragraph{Sentinel Hypothesis}
many investigators speculate that this central neural network reflects the brain’s relentless tracking and prediction of environmental events to adaptively optimize the organism's future action (Schacter et al., 2007).

Cognitive neuroimaging studies linked the DMN to the contemplation of others’ (Mar, 2011) and one’s thoughts (Lombardo et al., 2009) mental states,

DMN does social cognition -> but it exists in rats and monkeys with much more poorly developped social capacities
-> our computational framework can be equally applied to humans
and other primates.

@@

\section{Conclusion}
What single brain function could be most important for existence and survival of the species?
Estimating quantities that depend on how features of an agent’s environment are expected to unfold over the future.
From the perspective of the human condition, it remains unclear what the nature and purpose of a domain-overarching computational mechanism implemented in the high association cortex could be to warriant the high energetic costs.
MDP conceptualize control of the environment.
Obtain reward and avoid pain by adaptive behavior.

The idealized computational framework explains previous observations about the
DMN in a simple in conception but non-trivially links
together disparate functional hypotheses in the DMN literature.
The default brain function may be reducible to a small number
of well-defined equations.
biological plausibility of dynamic optimization in
optimal substructure
mathematical formalism; statistical solution; adaptive fitness
; formalize and predict; offer statistical control on the brain's default function/operations
stochastic modelling
At the least, we propose an alternative vocabulary to
describe and discuss neuroscientific studies on the DMN.
%
In principle,
neuroscientific experiments can be designed that operationalize
the set of action, value, and state variables that determine
the behavior of RL systems. This makes proposed machine-learning
approach to DMN function not only practically computable but
also neuroscientifically falsifiable.
%
The DMN can be viewed as a smoothing kernel extending
from the relatively recent past to the relatively close future
that constantly estimates complex non-linear gain functions.
-> This is why neuroimaging research revealed activity increases in the DMN
when 
-> This may be why perfect memory has been evolutionarily de-selected,
although it is provably feasible by the neurobiological hardware in humans

>>>>

The vmPFC subserves less ambiguous subjective-value-related evaluative processes
reward-informed and risk estimation of self-relevant environmental stimuli.
This DMN node is more closely associated with
orchestrating adapted behavior by bottom-up-driven
processing of “what matters now”,
probably drawing on model-based value representations
\cite{doherty2015structure}.
Quantative lesion findings across 344 human individuals confirmed
a gross impairment in value-based decision making
(Glaescher et al., 2016).
The vmPFC is preferentially connected with limbic and reward-related areas.
The vmPFC has been observed to have monosynaptical connections
with the nucleus accumbens
in axonal tracing studies in monkeys (Haber et al., 1995; Ferry et al., 2000).
The NAc is thought to be linked to reward mechanisms that not
only modulate motivated behavior towards basic survival needs but also
subserve reinforcement learning in humans more broadly
(O'Doherty 2015).
This is consistent with diffusion MRI (dMRI) tractography in humans and monkeys
(Croxson et al., 2005) that
quantified this brain regions to
be substantially more likely connected to the vmPFC than dmPFC in both species.
Two functional connectivity modalities in humans strongly connected
the vmPFC with the nucleus accumbens, hippocampus (HC),
and posteriomedial cortex.
%
The vmPFC is often proposed to be involved in (external) emotional
reactions and own (visceral) arousal.
Real or imaginged bodily states could be mapped in the vmPFC
as a bioregulatory disposition governing cognition
and decision making (Damasio, 1996; Nauta, 1971)
In neuroeconomic studies of human decision making,
the vmPFC consistently reflects an individual’s subjective
valuation (Behrens et al., 2008; Boorman et al., 2009; Knutson et al., 2005; Plassmann et al., 2007)
This may be why performance within and across participants
was related to state encoding in OFC.
Such a \textit{cognitive map} of action space was argued to encode
the current task state even when states are unobservable from sensory input,
which was shown to be critical for behavior.
The GMV of the vmPFC and nucleus accumbens
correlated with indices of social reward attitudes and
value-guided behavior (Lebreton et al., 2009).
Additionally,
independent whole-brain analyses from structural
neuroimaging studies related the gray-matter volume GMV of the vmPFC,
more consistently than any other
brain region, to indices of
social competence and social network complexity,
among the most complicated decision that humans and monkeys take
(Lebreton et al., 2009; Powell et al., 2010; Lewis et al., 2011; Sallet et al., 2011).
%

@@

By construction, in general, maximizing the expectation in  \eqref{eq:qval} is intractable due to
the enormous number of possible trajectories the agent might need to
consider.
(If the Bellman-based computational ideas have a corresponding
neurobiological implementation, this provides a framework of how
optimal behavior could be computable in the human brain.)
One popular simplification that removes this issue is to only consider the sub-family of
\textit{deterministic policies}, which map each state to the best single best action which
should be taken at thin state, and therefore define functions from states to actions.
\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The celebrated \textit{Bellman's equation} [Sutton and Barto 1998] then takes the simple form
\begin{equation}
  Q^\pi(s, a) = \mathbb E_{s' \sim \rho(.|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q^\pi(s', a')]
  = \mathbb E_{s' \sim \rho(.|s,a)} [r(s,a) + \gamma Q^\pi(s', \pi(s'))]
  \label{eq:bellman}
\end{equation}
which provides a recursive decomposition of optimal behavior.
A complicated dynamic programming optimization can be
broken into simpler subproblems at different time points,
which has often been closely related to the medial prefrontal part of the DMN
\cite{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value,
to guide action (i.e., improve the policy) towards a better state.
It is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

@@

\section{Relation to other cognitive models}
\subsection{Predictive Coding and Bayesian Brain Hypothesis}
The predictive coding and Bayesian brain framework 
is a frequently evoked idea related to the default mode function
\cite{bar2007}.
According to this view,
cortical responses emerge from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
This dynamic intercourse enables inference on the world by reconciling
gaps between fresh sensory input and stored prior information. 
Feed-forward sensory experience is constantly calibrated by
top-down modulation at various hierarchical processing levels.
back projections probably outnumber by far the existing input projection in the brain
\cite{salin1995corticocortical}.
It can explains the constructive, generative nature of sensory perception
\cite{friston2010free} and
why motor action is intimately related to sensory expectations
\cite{wolpert1995internal}.
At each stage of neural processing,
an internally generated prediction of the sensory input is
directly compared against the actual input from the environment.
A prediction error at one of the processing levels,
i.e. the difference between sensory input
and internally predicted sensation,
incurs plasticity changes of neuronal back-projections (i.e., model parameters)
for gradually improved future prediction of the environment.
Contextual integration is maintained by top-down modulation of sensorimotor
processing by context-specific a-priori information.
The generative model of how data arise in the
world hence be incorporated into
the current neuronal wiring.
Indeed,
This process permits updates of the internal model of the environment
to best fit the incoming sensory examples.

@@

\subsection{Semantic Hypothesis}
Another frequently proposed account of DMN function revolves
around forming meaning associations and abstract analogies between
the currently experienced environment and
stored knowledge from the past
\cite{bar2007proactive, binder1999conceptual}.
Analogies might naturally tie incoming novel sensory stimuli to
explicit world knowledge stored in past memories
\cite{bar2009proactive}.
Using these semantic analogies to mentally envision scenarios
would improve optimal behavior in the environment
integrated as new memories for future analogical associations.
Such cognitive processes can afford
the internal generation of necessary information
that is not presented in the surrounding environment.
This accounts further explains the reciprocal relationship
between an egocentric first person perspective and
an allocentric bird’s eye perspective immersed in
self-reflection, social knowledge, and autobiographical memories.
%
Closely related to episodic memory, language, problem solving,
planning, others' thoughts, and spatial navigation
as these necessitate abstract world knowledge and abstract assocations
for binding the constituent elements in mental scene construction
\cite{schacter2007remembering}.
These scene generation processes could contribute to interpreting the
present and foretelling the future.
This view is supported by evidence in animals that
could learn a \textit{cognitive map of the environment},
even without reward incentives, and exploit it later
for other means \cite{tolman1948cognitive}.

\begin{itemize}
  \item The notion of semantic or knowledge association is
  incorporated into the MDP as a function of the state and action
  history.
  \item Both: there is no evidence to indicate that predictions of various
  levels of complexity, abstraction, timescale and purpose
  use mechanisms that are qualitatively different. This concurs with
  DMN activity increases across time, space, and content domains
  demonstrated in many neuroimaging studies. The semantic hypothesis
  and RL account provide explanations why hippocampus lesion does
  not only impair retrieving memories, but also hypothetical and future
  thinking \cite{hassabis2007patients}.
  \item The semantic hypothesis and MDP expose mechanisms of
  how alternative decisions could be mentally explored.
  \item While mental scene construction entails a distinction between
  environmental perception and internally generated mind-wandering
  \footnote{``A computational model of how such a process might be structured
  is far from being defined, but it will probably require a form of
  regulation by which perception of the current world is suppressed
  while simulation of possible alternatives are constructed,
  followed by a return to perception of the present.''\cite{buckner2007self}},
  the MDP naturally integrates the former egocentric
  (related to the current action, state and reward) and the later
  allocentric (related to past and future actions, states, and rewards)
  angles on the world in a same optimization problem.
  \item The semantic account of DMN function does not offer 
  a clear explanation how semantic world knowledge and analogies thereof
  lead to prediction about future actions and states.
  \item While both semantic and MDP account propose memory-based internally
  generated information for probabilistic mental models of action outcomes,
  only MDPs render explicit the grounds on which the final action is
  actually chosen (i.e., the estimated cumulative reward).
  \item In sum, the semantic hypothesis lacks an explicit time and
  incentive model.
\end{itemize}

\subsection{Sentinel Hypothesis}
many investigators speculate that this central neural network reflects the brain’s relentless tracking and prediction of environmental events to adaptively optimize the organism's future action (Schacter et al., 2007).

Cognitive neuroimaging studies linked the DMN to the contemplation of others’ (Mar, 2011) and one’s thoughts (Lombardo et al., 2009) mental states, 
Schilbach et al., 2008
Clan mentality
DMN does social cognition -> but it exists in rats and monkeys with much more poorly developped social capacities
-> our computational framework can be equally applied to humans
and other primates.

@@

\section{Conclusion}
What single brain function could be most important for existence and survival of the species?
Estimating quantities that depend on how features of an agent’s environment are expected to unfold over the future.
From the perspective of the human condition, it remains unclear what the nature and purpose of a domain-overarching computational mechanism implemented in the high association cortex could be to warriant the high energetic costs.
MDP conceptualize control of the environment.
Obtain reward and avoid pain by adaptive behavior.
How representations of past and future are integrated into
action choice in the present context.
The idealized computational framework explains previous observations about the
DMN in a simple in conception but non-trivially links
together disparate functional hypotheses in the DMN literature.
The default brain function may be reducible to a small number
of well-defined equations.
biological plausibility of dynamic optimization in
optimal substructure
mathematical formalism; statistical solution; adaptive fitness
; formalize and predict; offer statistical control on the brain's default function/operations
stochastic modelling
At the least, we propose an alternative vocabulary to
describe and discuss neuroscientific studies on the DMN.
%
In principle,
neuroscientific experiments can be designed that operationalize
the set of action, value, and state variables that determine
the behavior of RL systems. This makes proposed machine-learning
approach to DMN function not only practically computable but
also neuroscientifically falsifiable.
%
The DMN can be viewed as a smoothing kernel extending
from the relatively recent past to the relatively close future
that constantly estimates complex non-linear gain functions.
-> This is why neuroimaging research revealed activity increases in the DMN
when 
-> This may be why perfect memory has been evolutionarily de-selected,
although it is provably feasible by the neurobiological hardware in humans
